Preschoolers form conventional pacts with each other to communicate about novel shapes

### 2025-01-27 write-ups/kid_tangrams_cogsci_2025/kid_tangrams_cogsci_2025.Rmd
Feedback + Edits

### Procedure
Once a pair of children agreed to play the game, a research assistant took them to a quiet testing room. Children were introduced to a stuffed animal "Smurfy" who wanted to play a matching game. Children sat across a table from each other, each with a tablet in front of them (Figure \ref{game}A). On each trial, one child was the "teller" and saw a black box around one of two images on their screen and was asked to "say what they saw" in the black box. The "guesser" saw the same two images in a randomized order and tried to select the described image to help Smurfy make a match. When the guesser selected an image, both children received feedback in form of a smiley or frowny face and an audible sound. After each trial, children switched roles. Children passed Smurfy back and forth to help them keep track of whether they were the "guesser" or "teller" on a given trial.

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children generally accurate in their selections. To test for changes in accuracy over time, we fit a Bayesian mixed effects logistic regression predicting accuracy.^[`r acc_form |> form()`] Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`), and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`). This level of accuracy is generally in-line with accuracies from 4-year-olds playing with their parents in @leung2024 and indicates that children can understand and succeed at the task.

@@

While description length is used as a proxy for measuring convention formation, semantic overlap between utterances does not always track description length. @boyce2024 introduced a more sensitive measure of semantic convergence.  By comparing the content of utterances using word embeddings, @boyce2024 traced how similarities within and across games changed over time. <!--To measure convergence of utterances, they compared how the similarities between utterances from a given block and the final block (conventionalized) utterances increased over the course of the game. -->

With only 3 blocks of descriptions, we do not believe it is meaningful to look at whether the 2nd and 3rd blocks are more similar than the 1st and 3rd blocks. However, we can test for a more coarse measure of sensitivity to partner: whether children's utterances are more like their partner's than children in other games'. If children are fully ego-centric [as suggested by @glucksberg1966], their choices of descriptions would be independent from their partners.

@@

We compared the semantic similarities between descriptions of the same target based on who produced the description (Figure \ref{sim}).  We used a Bayesian mixed effects linear regression to predict similarity.^[`r sim_across_between_form |> form()`] Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`). Utterances were slightly more similar still if they came from the same person within the partnership (`r sim_across_between_1_results |> stats_text(3,3)`), which is expected since children are likely to be fairly consistent with themselves. However, children used descriptions that were much more similar to their partner's than to other children's (Figure \ref{sim}), indicating some sensitivity to their partner's expressions.

@@

As a course measure of partner-sensitivity, we repeated the semantic analysis from Experiment 1. Utterances were more similar if they came from the same partnership (`r sim_across_between_2_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_2_results |> stats_text(3,3)`).

@@

We might expect that whether a description was successful influences whether the same description, or a variant of it, is employed in future rounds. Intuitively, successful descriptions can be copied and built upon, while unsuccessful descriptions should be replaced by a fresh attempt. To test this idea that accuracy might be predictive of similarity to future descriptions, we ran a post-hoc Bayesian linear model predicting similarity to the next block description in terms of accuracy.^[`r form(sim_acc_to_next_form)`] Descriptions that elicited a correct response were more similar with the next block description (`r stats_text(sim_acc_to_next_mod_meta, 2, 3)`) with no substantial interaction with block number or whether both descriptions came from the same teller. This pattern of results is consistent with children sticking to their own description more when it was previously successful last time, and children repeat the other child's description more when they understood it. <!--Numerically, the size of the boost for being correct was similar to the size of the boost from coming from the same person `r stats_text(sim_acc_to_next_mod_meta, 6,3)`. -->

@@

Prominent classical findings have been interpreted as evidence that young children are cannot overcome their ego-centrism to coordinate with each other in reference games [@glucksberg1966]. However, an emerging pattern of research finds that  young children show emerging communicative and pragmatic sensitivity, especially when cognitive demands are low. <!--Recent work has established that children are successful at producing and understanding referential expressions to novel shapes when playing reference games with their parents. -->We re-addressed the question of preschoolers's performance in child-child reference games, using a scaffolded paradigm to reduce extraneous task demands.

Across 2 experiments and 51 pairs of 4 and 5-year old children, we tested how well children could produce referential expressions that allowed their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (85%), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts. Children did not display a reduction in the length of referential expressions over the course of the game. While this task is substantially scaled down relative to measures used for adult competence, it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

It is unclear to what extent the uniformly short descriptions are a product of the simplified task or children's behavioral differences from adults. In this case, the low number of options and relatively easy to describe shapes may have obviated the need for long initial descriptions. Indeed, adults controls in @leung2024 used shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images [@hawkins2020b; @boyce2024]. However, young children may also struggle to produce longer descriptions, and young children may be more willing to take guesses when adults would seek additional clarification. Especially in light of other work suggesting that conceptual pact formation and reduction in utterance length sometimes decouple in adults [@boyce2024], further empirical work on the factors driving verbosity in reference games is warranted.

The generalizability of our results is limited by the target population, the target images, and the task structure. We sampled a convenience poplaution of children at a university nursery school, who are far from representative of all 4 and 5 year olds. The set of tangram images may be easier to distinguish have higher codability than other target images used in adult reference games. We specifically targeted children's abilities to construct referring expressions that can be jointly understood, so children were provided scaffolding around the larger coordination problems of taking turns and talking to their partner. Thus, children's performance should be taken as a proof-of-concept about ability, rather than a claim about how generally children demonstrate these abilities.

In the broader picture of language acquisition, there's debate over the timing of emergence of communicative and pragmatic abilities relative to the acquisition of grammar and meaning. On one side, children seem to learn literal semantics far before they display an understanding of some pragmatic implicatures [@huang2009, @noveck2001]; on the other, sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning [@bates1974; @bohn2019a; @tomasello2008]. The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cognitive capacities increase. At 4–5 years old, even though their utterances are not fully adult-like in form, children are nonetheless able to use their linguistic skills communicatively.

>>>>

### Procedure
Once a pair of children agreed to play the game, a research assistant took them to a quiet testing room. Children were introduced to a stuffed animal "Smurfy" who wanted to play a matching game. Children sat across a table from each other, each with a tablet in front of them (Figure \ref{game}A). On each trial, one child was the "teller" and saw a black box around one of two images on their screen and was asked to "tell Smurfy what they see" in the black box. The "guesser" saw the same two images in a randomized order and tried to select the described image to help Smurfy make a match. When the guesser selected an image, both children received feedback in form of a smiley or frowny face and an audible sound. After each trial, children switched roles. Children passed Smurfy back and forth to help them keep track of whether they were the "guesser" or "teller" on a given trial.

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children were generally accurate in their selections. To test for changes in accuracy over time, we fit a Bayesian mixed effects logistic regression predicting accuracy.^[`r acc_form |> form()`] Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`), and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`). This level of accuracy is generally in-line with accuracies from 4-year-olds playing with their parents in @leung2024 and indicates that children can understand and succeed at the task.

@@

While description length is often used as a proxy for measuring convention formation, it does not capture semantic overlap between utterances. @boyce2024 introduced a more sensitive measure of semantic convergence that compares the content of utterances using word embeddings to trace how similarities within and across games change over time. <!--To measure convergence of utterances, they compared how the similarities between utterances from a given block and the final block (conventionalized) utterances increased over the course of the game. --> With only 3 blocks of descriptions, we do not expect semantic similarity for descriptions of a given target to show any meaningful change over time. However, we can test for a more coarse measure of sensitivity to partner: whether children's utterances are more like their partner's than children in other games'. If children are fully ego-centric [as suggested by @glucksberg1966], their choices of descriptions would be independent from their partners.

@@

We compared the semantic similarities between descriptions of the same target based on who produced the description (Figure \ref{sim}).  We used a Bayesian mixed effects linear regression to predict similarity.^[`r sim_across_between_form |> form()`] Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`). Utterances were slightly more similar still if they came from the same person within the partnership (`r sim_across_between_1_results |> stats_text(3,3)`), which is expected since children are likely to be fairly consistent with themselves. However, children used descriptions that were much more similar to their partner's than to other children's (Figure \ref{sim}), indicating sensitivity to their partner's expressions.

@@

As a course measure of partner-sensitivity, we repeated the semantic analysis from Experiment 1. Utterances were more similar if they came from the same partnership (`r sim_across_between_2_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person within the partnership (`r sim_across_between_2_results |> stats_text(3,3)`).

@@

We might expect that whether a description was successful influences whether the same description, or a variant of it, is employed in future rounds. Intuitively, successful descriptions can be copied and built upon, while unsuccessful descriptions should be replaced by a fresh attempt. To test this idea that accuracy might be predictive of similarity to future descriptions, we ran a post-hoc Bayesian linear model predicting similarity to the next block description in terms of accuracy.^[`r form(sim_acc_to_next_form)`] Descriptions that elicited a correct response were more similar with the next block description (`r stats_text(sim_acc_to_next_mod_meta, 2, 3)`) with no substantial interaction with block number or whether both descriptions came from the same teller. This pattern of results is consistent with the expectation that children are more likely to stick to their own description or repeat the other child's description if it was previously successful. <!--Numerically, the size of the boost for being correct was similar to the size of the boost from coming from the same person `r stats_text(sim_acc_to_next_mod_meta, 6,3)`. -->

@@

Prominent classical findings have been interpreted as evidence that young children cannot overcome their ego-centrism to coordinate with each other in reference games [@glucksberg1966]. However, an emerging pattern of research finds that  young children show emerging communicative and pragmatic sensitivity, especially when cognitive demands are low. <!--Recent work has established that children are successful at producing and understanding referential expressions to novel shapes when playing reference games with their parents. -->We re-addressed the question of preschoolers's performance in child-child reference games, using a scaffolded paradigm to reduce extraneous task demands.

Across 2 experiments and 51 pairs of 4 and 5-year old children, we tested how well children could produce referential expressions that allowed their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (85%), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts. While this task is substantially scaled down relative to measures used for adult competence, it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

Unlike adults, Children did not display an increase in accuracy nor a reduction in the length of referential expressions over the course of the game. This might be taken to mean that children are not adapting their descriptions over time. Still, our findings show that descriptions became increasingly similar to descriptions in the last block and that successful utterances were more similar to future utterances, suggesting that childen are adapting their descriptions as the game unfolds. These null findings are likely a result of initially high accuracy in the first block and initially short utterances that have less possibility for reduction when compared to adult utterances.

It is unclear to what extent the uniformly short descriptions are a product of the simplified task or children's behavioral differences from adults. In this case, the low number of options and relatively easy to describe shapes may have obviated the need for long initial descriptions. Indeed, adult controls in @leung2024 used shorter initial descriptions than adults in studies with larger arrays of harder to distinguish images [@hawkins2020b; @boyce2024]. However, young children may also struggle to produce longer descriptions, and young children may be more willing to take guesses when adults would seek additional clarification. Especially in light of other work suggesting that conceptual pact formation and reduction in utterance length sometimes decouple in adults [@boyce2024], further empirical work on the factors driving verbosity in reference games is warranted.

The generalizability of our results is limited by the target population, the target images, and the task structure. We sampled a convenience poplaution of children at a university nursery school, who are far from representative of all 4 and 5-year olds. The set of tangram images may be easier to distinguish and have higher codability than other target images used in adult reference games. We specifically targeted children's abilities to construct referring expressions that can be jointly understood, so children were provided scaffolding around the larger coordination problems of taking turns and talking to their partner. Thus, children's performance should be taken as a proof-of-concept about ability, rather than a claim about how generally children spontaneously demonstrate these abilities.

In the broader picture of language acquisition, there's debate over the timing of emergence of communicative and pragmatic abilities relative to the acquisition of grammar and meaning. On one side, children seem to learn literal semantics far before they display an understanding of some pragmatic implicatures [@huang2009, @noveck2001]; on the other, sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning [@bates1974; @bohn2019a; @tomasello2008]. The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cognitive capacities increase. At 4–5-years old, even though their utterances are not fully adult-like in form, children are nonetheless able to use their linguistic skills communicatively.
